Verifying Cache Coherence Protocols for an Infinite Family of Cache Hierarchies

### 2014-02-06 paper/Intuition.tex
A pass through Intuition

\section{Proof of MSI protocol being preserving Store atomicity}
\label{safety}

@@

We first reduce the proof of store atomicity to 4 invariants, presented in
Section \ref{firstLevel}. These invariants in turn can be reduced to 4 more
invariants presented in Section \ref{secondLevel}. The reason for choosing the 8
 invariants is because they are not specific to MSI protocol
alone; any other protocol which obeys these invariants obeys Theorem
\fullref{storeAtomicity}. We finally present the key insight which we
discovered for proving the second-level invariants in the form of 3 invariants.
We believe that other invalidation-based protocols with inclusive caches (\ie{}
those cache hierarchies where a parent cache contains all the locations that
any of its children contains) like MESI, MOSI and MOESI also
satisfy these 3 fundamental properties, making their verification easier by reusing
portions of our MSI proof.
We informally show why this is the case using the example of MESI,
MOSI and MOESI in Section \ref{mosi}.

@@

These invariants hold after any number of executions of the transactions starting
from the initial state. We use the term time $t$ to specify the global state
reached after firing $t$ transitions.

While we have proofs in Coq for each invariant our discussion will be mainly
informal for brevity. A full description can be found at~\xxx{MURALI PUT
REFERENCE HERE}.

\subsection{First level invariants}
\label{firstLevel}
Intuitively our protocol is correct because it guarantees that at most one
cache can handle a store request at a time, and that any cache that attempts to
handle a load request has the ``latest'' value. We can state this formally with
the following 4 invariants.

\begin{inv}
\textit{latestValue}:
At any time $t$, a cache $c$ which is clean for an address $a$ will have the most
up-to-date value for that address, \ie{}
\begin{enumerate}
\item $c.\data[a] = \initData(a)$ and no store request $q'$ has been processed at
any time $t'$ such that $t' < t \wedge a = q'.\addrQ$, or
\item There is a store request $q$ that was processed at time $t_q$ such that
$t_q < t \wedge a = q.\addrQ \wedge c.\data[a] = q.\dataQ$ and no other store
request $q'$ was processed at any time $t'$ such that $t_q < t' < t \wedge a =
q'.\addrQ$.
\end{enumerate}
\label{latestValue}
\end{inv}
A clean cache is defined as follows.
\begin{defn}
\textit{clean}: A cache is said to be \clean{} for an address $a$ if and only if
the state of the cache for that address is at least \Sh{} and the directory 
state for all its children is at most \Sh.
\label{clean}
\end{defn}

\begin{inv}
\textit{nonAncestorCompatible}: For two distinct caches $c_1$ and $c_2$ such
that neither is an ancestor of the other in the cache tree hierarchy (where
ancestor relation is the reflexive-transitive closure of the parent relation),
for each address $a$, $c_1.\state[a]$ and $c_2.\state[a]$ are
\textit{compatible}.
\label{nonAncestorCompatible}
\end{inv}
Compatibility of states is defined as follows.
\begin{defn}
\textit{compatibile}: Values of states $x_1$ and $x_2$ are \textit{compatible}
iff either $x_1$ or $x_2$ is \Mo{}, then the other is \In.
\label{compatible}
\end{defn}

It is easy to see that any two states $y_1 \le x_1$ and $y_2 \le x_2$ are also
compatible if $x_1, x_2$ are compatible.

\begin{inv}
\textit{processReq}: Load and store requests can be processed only at a leaf
node which is at least in the \Sh{} and \Mo{} states, respectively, for that
address.
\label{processReq}
\end{inv}

\begin{inv}
\textit{noReqAgain}:
\item For a leaf cache $c$, $c.\pos$ changes only on processing a request. In
that case, $c.\pos$ gets incremented.
\label{noReqAgain}
\end{inv}

Given these invariants, we will show how Theorem~\fullref{storeAtomicity}
follows.
%From the transitions of Figure \ref{trans}, a store may happen only
%when a leaf cache is in \Mo{} state for that
%address (Invariant~\textit{processReq}).
Leaf caches cannot be ancestors of any other node, so by Invariant
\fullref{nonAncestorCompatible}, if a leaf cache is in \Mo{} state, all other
leaf caches will be in \In{} state preventing the possibility of processing
load and store requests (because of Invariant \ref{processReq}) for the same
address at the same time at different leaf caches. This satisfies
Definition~\fullref{uniqStTimes}.

It is easy to see that Definition \fullref{uniqRespLabels}, Definition
\fullref{localOrdering} and Definition \fullref{allPrevious} directly follows
from Invariant \fullref{noReqAgain}.  As leaf caches have no children, one in
at least \Sh{} state for address $a$ is \clean{} for that address. Since
processing a load request requires the cache to be at least in the \Sh{} state,
its response, which supplies the data for that address present in the cache,
supplies the latest value by Invariant~\fullref{latestValue}. Thus, all the
conditions for Theorem~\fullref{storeAtomicity} is satisfied given Invariants
\ref{latestValue} to \ref{noReqAgain}.

\subsection{Second-level Invariants}
\label{secondLevel}

While Invariant \fullref{processReq} and \fullref{noReqAgain} can be seen
directly from the MSI transitions, correctness of Invariants
\fullref{nonAncestorCompatible} and \fullref{latestValue} require further
decomposition. We present 4 more invariants for proving those two invariants.

@@

\begin{inv}
\textit{noTransitWrite}: Whenever data for an address $a$ is in transit (\ie{}
$\forall t, t_s \le t \le t_r$ where $t_s$ is the time of sending the data and
$t_r$ the time of receiving the data), no cache can process a store request for
$a$, and the data must be sent from a cache which is \textit{clean}.
\label{noTransitWrite}
\end{inv}

@@

Any two caches $c_1, c_2$ neither of which are an ancestor of the other has a
lowest common ancestor $x$ which is an ancestor of both $c_1$ and $c_2$. $x$
has two distinct children $x_1$ and $x_2$ such that they are the ancestors of
$c_1$ and $c_2$ respectively. For any address $a$, $x_1.\state[a]$ and
$x_2.\state[a]$ are compatible by 
Invariant~\fullref{localCompatibility2}. It follows that $x_1.\state[a]$ and
$x_2.\state[a]$ are compatible since these values are smaller than the
respective directory values. From the previous paragraph about the relation
between the states of a cache and its ancestor, it follows that $c_1$ and $c_2$
are also compatible. This proves Invariant \fullref{nonAncestorCompatible}.

We prove Invariant \fullref{latestValue} by strong induction on time. Consider
two cases: a) cache $c$ is already \clean{} for an address $a$ at time $t$ or
b) the contrary.  If \clean{}, then it has the latest value at time $t$. Since
none of its children's directory entries can be in \Mo{}, all its descendants
should be no greater than \Sh{} for that address (because of Invariant
\fullref{conservative} and the fact that $\ge$ relation is preserved for the
state of a cache and that of its ancestor). Any non-ancestor of $c$, and
non-descendant of $c$ must be compatible (as discussed in the previous
paragraph).  Since $c$ is \clean, it is at least in the \Sh{} state, so
compatibility means that non-descendants can not be in \Mo{} state. All of
$c$'s ancestors are non-leaves, so they can not process any store request.
Therefore, no cache save $c$ may have process a store request, which means that
$c$ contains the latest value for address $a$ at time $t+1$ if it was \clean{}
at time $t$.

If $c$ was not clean at time $t$, it must have received data from either its
parent or its child. And the sender is \clean{} by Invariant
\ref{noTransitWrite}. By the induction hypothesis, it means that the sender has
the latest value for the address. Again, because of
Invariant~\ref{noTransitWrite}, no store request can be processed when the data
is in transit, which means $c$ will have the latest value for address $a$ at
time $t+1$. This establishes Invariant \fullref{latestValue}.

\subsection{Key insight for proving second-level invariants}

%ndave: This needs some repointing
We present what we believe to be the key insight for proving any cache coherence protocol.
In fact the request/response message semantics for the MSI protocol was designed to satisfy
the following three invariants.

\begin{inv}
\textit{stateChange}:
\begin{enumerate}
\item A cache downgrades its state for an address iff it sends a response for that address
\item A parent upgrades the directory entry for an address for one of its
children iff it sends a response for that child and address
\item The state or directory entry changes for an address only on sending and
receiving an response for that address, and they change to the \myto-field of the response.
\end{enumerate}
\label{stateChange}
\end{inv}

@@

\begin{figure}
\centering
\includegraphics[scale=.5]{resps}
\caption{An example showing a sequence of responses ping ponging between a child and its parent}
\label{resps}
\end{figure}

Figure \ref{resps} shows an example of the effect of Invariants
\fullref{stateChange}, \fullref{noCross} and \fullref{respFIFO}. The directory
entry of the parent or the state of the child changes on sending and receiving
a response and the new state is showed at the points of sending or receiving
responses.
Both the cache and its parent follow the same sequence state and directory
entry changes; just that the receiver always lags slightly behind the sender.
The receiver can not send any response before receiving because of
\fullref{noCross}, thereby eliminating any scenarios where the lagging makes
reasoning complicated.

Using these three invariants (Invariant \ref{stateChange} to \ref{respFIFO}, it
is easy to see that Invariant \fullref{conservative} holds. Intuitively, going
back to Figure \ref{resps}, we can see that at all times, either there is no
response in flight, or only responses from parent in flight, or responses from
child in flight. A response from child downgrades the state and a response from
parent upgrades the directory. Since the other side simply lags behind,
Invariant \fullref{conservative} holds. Since Invariants \ref{stateChange} to
\ref{respFIFO} deals with just one cache's state for a single address and the
corresponding directory entry, these properties can potentially be verified
using model-checking techniques for an invalidation-based protocol.

Proving the other 3 invariants in Section \ref{secondLevel} requires using
more properties of the MSI transistions of Figure \ref{trans}, though it is
straightforward. Analysing a parent-child pair will suffice in proving Invariants
\fullref{localCompatibility1} and \fullref{localCompatibility2}. Proof of Invariant
\fullref{noTransitWrite} is similar to the proof of Invariant
\fullref{nonAncestorCompatible}.

>>>>

\section{Proof that the MSI Protocol Preserves Store Atomicity}
\label{safety}

@@

We first reduce the proof of store atomicity to 4 invariants, presented in
Section \ref{firstLevel}. These invariants in turn can be reduced to 4 more
invariants presented in Section \ref{secondLevel}. The reason for choosing the 8
 invariants is because they are not specific to the MSI protocol;
any other protocol that obeys these invariants obeys Theorem
\fullref{storeAtomicity}. We finally present the key insight that we
discovered for proving the second-level invariants in the form of 3 invariants.
We believe that other invalidation-based protocols with inclusive caches (\ie{}
those cache hierarchies where a parent cache contains all the locations that
any of its children contains) like MESI, MOSI and MOESI also
satisfy these 3 fundamental properties, making their verification easier by reusing
portions of our MSI proof.
We informally show why this is the case using the examples of MESI,
MOSI and MOESI in Section \ref{mosi}.
\adamc{...and maybe we don't, for space reasons. \texttt{:-)}}

@@

These invariants hold after any number of executions of the transitions starting
from the initial state. We write ``time $t$'' for the global state
reached after firing $t$ transitions.

While we have proofs in Coq for each invariant, our discussion will be mainly
informal for brevity. A full description can be found at~\xxx{MURALI PUT
REFERENCE HERE}. \adamc{or just remind of the source code URL given earlier}

\subsection{First-Level Invariants}
\label{firstLevel}
Intuitively our protocol is correct because it guarantees that at most one
cache can handle a store request at a time, and that any cache that attempts to
handle a load request has the ``latest'' value. Four invariants formalize
the intuition.

\begin{inv}
\textit{latestValue}:
At any time $t$, a cache $c$ that is clean for an address $a$ will have the most
up-to-date value for that address, \ie{}
\begin{enumerate}
\item $c.\data[a] = \initData(a)$ and no store request $q'$ has been processed at
any time $t'$ such that $t' < t \wedge a = q'.\addrQ$, or
\item There is a store request $q$ that was processed at time $t_q$ such that
$t_q < t \wedge a = q.\addrQ \wedge c.\data[a] = q.\dataQ$ and no other store
request $q'$ was processed at any time $t'$ such that $t_q < t' < t \wedge a =
q'.\addrQ$.
\end{enumerate}
\label{latestValue}
\end{inv}

\vspace{-.15in}

A clean cache is defined as follows.
\begin{defn}
\textit{clean}: A cache is said to be \clean{} for an address $a$ if and only if
the state of the cache for that address is at least \Sh{} and the directory 
state for all its children is at most \Sh.
\label{clean}
\end{defn}

\begin{inv}
\textit{nonAncestorCompatible}: For two distinct caches $c_1$ and $c_2$ such
that neither is an ancestor of the other in the cache tree hierarchy (where
the ancestor relation is the reflexive-transitive closure of the parent relation),
for each address $a$, $c_1.\state[a]$ and $c_2.\state[a]$ are
\textit{compatible}.
\label{nonAncestorCompatible}
\end{inv}
Compatibility of states is defined as follows.
\begin{defn}
\textit{compatibile}: Values of states $x_1$ and $x_2$ are \textit{compatible}
iff whenever either $x_1$ or $x_2$ is \Mo{}, the other is \In.
\label{compatible}
\end{defn}

It is easy to see that any two states $y_1 \le x_1$ and $y_2 \le x_2$ are also
compatible if $x_1$ and $x_2$ are compatible.

\begin{inv}
\textit{processReq}: Load and store requests can be processed only at a leaf
node that is at least in the \Sh{} and \Mo{} states, respectively, for that
address.
\label{processReq}
\end{inv}

\begin{inv}
\textit{noReqAgain}: For a leaf cache $c$, $c.\pos$ changes only on processing a request. In
that case, $c.\pos$ is incremented.
\label{noReqAgain}
\end{inv}

Given these invariants, we will show how Theorem~\fullref{storeAtomicity}
follows.
%From the transitions of Figure \ref{trans}, a store may happen only
%when a leaf cache is in \Mo{} state for that
%address (Invariant~\textit{processReq}).
Leaf caches cannot be ancestors of any other node, so by Invariant
\fullref{nonAncestorCompatible}, if a leaf cache is in \Mo{} state, all other
leaf caches will be in \In{} state, blocking simultaneous
load and store requests (because of Invariant \ref{processReq}) for the same
address at the same time at different leaf caches. Thus, we have proved
Definition~\fullref{uniqStTimes}.

It is easy to see that Definition \fullref{uniqRespLabels}, Definition
\fullref{localOrdering}, and Definition \fullref{allPrevious} directly follow
from Invariant \fullref{noReqAgain}.  As leaf caches have no children, one in
at least \Sh{} state for address $a$ is \clean{} for that address. Since
processing a load request requires the cache to be at least in the \Sh{} state,
its response, which supplies the data for that address present in the cache,
supplies the latest value by Invariant~\fullref{latestValue}. Thus, all the
conditions for Theorem~\fullref{storeAtomicity} are satisfied given Invariants
\ref{latestValue} to \ref{noReqAgain}.

These arguments are fleshed out in a machine-checked Coq proof that the invariants
imply store atomicity.  Our proof may be reused for any protocol (i.e., alternative
choice of rules like in Figure \ref{trans}) that satisfies the invariants,
so that the only new verification effort is in proving the invariants.

\subsection{Second-Level Invariants}
\label{secondLevel}

While Invariants \fullref{processReq} and \fullref{noReqAgain} can be seen
directly from the MSI transitions, correctness of Invariants
\fullref{nonAncestorCompatible} and \fullref{latestValue} requires further
decomposition. We present 4 more invariants that are helpful in the proof.

@@

\begin{inv}
\textit{noTransitWrite}: Whenever data for an address $a$ is in transit (\ie{}
$\forall t. \; t_s \le t \le t_r$ where $t_s$ is the time of sending the data and
$t_r$ the time of receiving the data), no cache can process a store request for
$a$, and the data must be sent from a \textit{clean} cache.
\label{noTransitWrite}
\end{inv}

@@

Any two caches $c_1, c_2$ neither of which is an ancestor of the other have a
lowest common ancestor $x$ that is an ancestor of both $c_1$ and $c_2$. $x$
has two distinct children $x_1$ and $x_2$ such that they are the ancestors of
$c_1$ and $c_2$ respectively. For any address $a$, $x_1.\state[a]$ and
$x_2.\state[a]$ are compatible by 
Invariant~\fullref{localCompatibility2}. It follows that $x_1.\state[a]$ and
$x_2.\state[a]$ are compatible since these values are smaller than the
respective directory values. From the previous paragraph about the relation
between the states of a cache and its ancestor, it follows that $c_1$ and $c_2$
are also compatible. This proves Invariant \fullref{nonAncestorCompatible}.

We prove Invariant \fullref{latestValue} by strong induction on time. Consider
two cases: (a) cache $c$ is already \clean{} for an address $a$ at time $t$ or
(b) the contrary.  If \clean{}, then it has the latest value at time $t$. Since
none of its children's directory entries can be in \Mo{}, all its descendants
should be no greater than \Sh{} for that address (because of Invariant
\fullref{conservative} and the fact that the $\ge$ relation is preserved for the
state of a cache and that of its ancestor). Any non-ancestor of $c$ and
non-descendant of $c$ must be compatible (as discussed in the previous
paragraph).  Since $c$ is \clean, it is at least in the \Sh{} state, so
compatibility means that non-descendants cannot be in \Mo{} state. All of
$c$'s ancestors are non-leaves, so they cannot process any store requests.
Therefore, no cache save $c$ may have processed a store request, which means that
$c$ contains the latest value for address $a$ at time $t+1$ if it was \clean{}
at time $t$.

If $c$ was not clean at time $t$, it must have received data from either its
parent or its child, and the sender is \clean{} by Invariant
\ref{noTransitWrite}. By the induction hypothesis, it means that the sender has
the latest value for the address. Again, because of
Invariant~\ref{noTransitWrite}, no store request can be processed when the data
is in transit, which means $c$ will have the latest value for address $a$ at
time $t+1$, establishing Invariant \fullref{latestValue}.

The full version of this proof is also implemented in Coq, providing an even
higher-level starting point for verification of a new protocol.

\subsection{Key Insight for Proving Second-Level Invariants}

%ndave: This needs some repointing
We present what we believe to be the key insight for proving any cache-coherence protocol.
In fact, the request/response message semantics for the MSI protocol was designed to satisfy
the following three invariants.

\begin{inv}
\textit{stateChange}:
\begin{enumerate}
\item A cache downgrades its state for an address iff it sends a response for that address.
\item A parent upgrades the directory entry for an address for one of its
children iff it sends a response for that child and address.
\item The state or directory entry changes for an address only on sending and
receiving a response for that address, and they change to the \myto-field of the response.
\end{enumerate}
\label{stateChange}
\end{inv}

\vspace{-.1in}

@@

\begin{figure}
\centering
\includegraphics[scale=.5]{resps}
\caption{An example showing a sequence of responses ping-ponging between a child and its parent
\adamc{For a camera-ready version, we should replace these bitmap images with vector images, for better printing.}}
\label{resps}
\end{figure}

Figure \ref{resps} shows an example of the effect of Invariants
\fullref{stateChange}, \fullref{noCross} and \fullref{respFIFO}. The directory
entry of the parent or the state of the child changes on sending and receiving
a response and the new state is shown at the points of sending or receiving
responses.
Both the cache and its parent follow the same sequence of state and directory
entry changes, but the receiver always lags slightly behind the sender.
The receiver cannot send any response before receiving because of
\fullref{noCross}, thereby eliminating any scenarios where the lag complicates
reasoning.

Using these three invariants (Invariants \ref{stateChange} to \ref{respFIFO}), it
is easy to see that Invariant \fullref{conservative} holds. Intuitively, going
back to Figure \ref{resps}, we can see that at all times, either there is no
response in flight, or only responses from parent in flight, or responses from
child in flight. A response from child downgrades the state, and a response from
parent upgrades the directory. Since the other side simply lags behind,
Invariant \fullref{conservative} holds. Since Invariants \ref{stateChange} to
\ref{respFIFO} deals with just one cache's state for a single address and the
corresponding directory entry, these properties can potentially be verified
using model-checking techniques for an invalidation-based protocol.

Proving the other 3 invariants in Section \ref{secondLevel} requires using
more properties of the MSI transistions of Figure \ref{trans}, though the argument is
straightforward. Analyzing a parent-child pair will suffice in proving Invariants
\fullref{localCompatibility1} and \fullref{localCompatibility2}. Proof of Invariant
\fullref{noTransitWrite} is similar to the proof of Invariant
\fullref{nonAncestorCompatible}.
